Effects of resonant and random excitations on the proton beam in the Large Hadron Collider, with applications to the design of pulsed hollow electron lenses for active halo control

### 2018-03-25 resex.tex
Minor changes in preparation for submission: PhySH keywords instead of PACS, kth command, preprint number, spelled out rms, no horizontal lines in damper on/off figure, comma after i.e.

\documentclass[
prstab
,reprint
%,draft
%,preprint,tightenlines
,linenumbers
,longbibliography
,preprintnumbers
,showpacs
,showkeys
,amsfonts,amssymb,amsmath
,floatfix
]{revtex4-1}

\usepackage{bm}
\usepackage{dcolumn}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage[colorlinks=true, allcolors=NavyBlue]{hyperref}

@@

% abbreviations
\newcommand{\kthtp}{$k$th-turn pulsing}
\newcommand{\seventhtp}{7th-turn pulsing}
\newcommand{\eighthtp}{8th-turn pulsing}
\newcommand{\tenthtp}{10th-turn pulsing}

@@

% PACS 2008:
% 29.20.db Storage rings and colliders

\pacs{29.20.D-}% PACS, the Physics and Astrtheonomy
                             % Classification Scheme.
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle

@@

Considering past, current and future high energy colliders, each new
machine has represented a considerable leap in stored beam energy
(Table~\ref{tab:stored_energy}). Furthermore, recent measurements at
the LHC show that the tails of the transverse beam distribution are
overpopulated compared to a Gaussian distribution. This results in a
considerable amount of energy being stored in the beam tails. In
particular, in the case of the LHC, about 5\% of the beam population
is stored in the tails (i.e., above 3.5$\sigma$, where $\sigma$ is the
Gaussian rms core beam size), compared to 0.22\% in an ideal Gaussian
distribution, leading to 19~MJ of stored energy for nominal LHC
parameters and 34~MJ in the case of
HL-LHC~\cite{helreview_valentino}. This leads to the conclusion that a
mechanism is needed to deplete the beam tails in a controlled
manner. Further information on the needs for halo control in LHC can
be found in Ref.~\cite{helreview}.

@@

\item \textbf{resonant excitation:} The electron beam is switched on
  only every $k$-th turn. The excitation can be represented
  by the following expression:
  \begin{eqnarray}
    \label{intro:eqn:1}
    f(t) & = & \sum_{n=-\infty}^{+\infty}\delta\left[t-n\cdot(kT)\right],
  \end{eqnarray}
  where $n$ is the turn number and $T$ is the revolution period. Its
  Fourier representation is
  \begin{equation}
    \label{intro:eqn:2}
    f(t) = \Sh_{kT}(t) = \frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt},
  \end{equation}
  with $f_n = n \cdot f_{\rm rev} / k$ and where $\Sh_{kT}$ is the
  Dirac comb. In general, $k$-th~turn pulsing drives $k$-th order
  resonances~\cite{md_sim_hel_res_ex_fitterer}. This type of pulsing
  pattern was used in the Tevatron during regular collider operations
  for abort-gap cleaning~\cite{hel_tevatron_abortgap_zhang}.
\end{itemize}

@@

\begin{figure*}
  \begin{tabular}{ccc}
    \hline
    \multicolumn{3}{c}{Transverse damper off} \\ \hline
    Random excitation, V (2017) & \seventhtp, V (2017) & \tenthtp, V
                                                         (2016) \\
    \includegraphics[width=\thirdwidth]{2017_emitv_avg_rel_vran_no_damper.png}&
    \includegraphics[width=\thirdwidth]{2017_emitv_avg_rel_v7th_no_damper_no_text.png} &
    \includegraphics[width=\thirdwidth]{2016_emitv_avg_rel_v10th_no_damper_no_text.png}\\
    \hline
    \multicolumn{3}{c}{Transverse damper active} \\ \hline
    Random excitation, V (2017) & \seventhtp, V (2017) & \tenthtp, V
                                                         (2016) \\
    \includegraphics[width=\thirdwidth]{2017_emitv_avg_rel_vran_with_damper.png} &
    \includegraphics[width=\thirdwidth]{2017_emitv_avg_rel_v7th_with_damper_no_text.png} &
    \includegraphics[width=\thirdwidth]{2016_emitv_avg_rel_v10th_with_damper_no_text.png} \\
    \end{tabular}
    \caption{Comparison of measured vertical emittances for bunches
      with transverse damper off (top row) and with transverse damper
      on (bottom row), during random V excitations (left), \seventhtp\
      in V (center) and \tenthtp\ in V (right). (Because of a change
      in experimental setup, for the bunches represented by the yellow
      line in the bottom right plot, the excitation was zero instead
      of the maximum value.)}
    \label{fig:damp}
\end{figure*}

@@

These studies were motivated in part by the need to assess the effects
of a hollow electron lens (HEL), which is being proposed for active
halo control in HL-LHC. Hollow electron beam collimation was
demonstrated in the Tevatron in continuous
mode~\cite{hel_tevatron_stancari}, i.e.\ with the same electron beam
current acting on a given bunch every turn. In this case, the unwanted
residual kicks, which arise from profile asymmetries, injection and
extraction bends, solenoid field misalignments, etc. have a negligible
effect on the circulating beam. An electron lens that is pulsed
resonantly can have much stronger effects, and this fact was used in
the Tevatron for abort-gap
cleaning~\cite{hel_tevatron_abortgap_zhang}. The resonant-pulsing
capability (having different currents every turn for a given set of
bunches) will be incorporated in the HL-LHC hollow electron lens to
enhance its range of achievable removal rates. However, due to time
constraints, only a very limited set of experimental tests with
resonant kicks could be done in the Tevatron. More studies are
currently being planned using the electron lenses in the Relativistic
Heavy Ion Collider (RHIC) at Brookhaven National
Laboratory~\cite{Fischer:PRL:2015}.

>>>>

\documentclass[
prstab
,reprint
%,draft
%,preprint,tightenlines
,linenumbers
,longbibliography
,preprintnumbers
%,showpacs
,showkeys
,amsfonts,amssymb,amsmath
,floatfix
]{revtex4-1}

\usepackage[american]{babel}
\usepackage{bm}
\usepackage{dcolumn}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage[colorlinks=true, allcolors=NavyBlue]{hyperref}

@@

% abbreviations
\newcommand{\kth}{$k$th}
\newcommand{\kthtp}{$k$th-turn pulsing}
\newcommand{\seventhtp}{7th-turn pulsing}
\newcommand{\eighthtp}{8th-turn pulsing}
\newcommand{\tenthtp}{10th-turn pulsing}

@@

%\pacs{}

\keywords{High-energy accelerators and colliders; Nonlinear beam
  dynamics; Beam resonances; Beam control; Collimation}

\preprint{FERMILAB-PUB-18-084-AD-APC}

\maketitle

@@

Considering past, current and future high energy colliders, each new
machine has represented a considerable leap in stored beam energy
(Table~\ref{tab:stored_energy}). Furthermore, recent measurements at
the LHC show that the tails of the transverse beam distribution are
overpopulated compared to a Gaussian distribution. This results in a
considerable amount of energy being stored in the beam tails. In
particular, in the case of the LHC, about 5\% of the beam population
is stored in the tails (i.e., above 3.5$\sigma$, where $\sigma$ is the
Gaussian root-mean-square beam size), compared to 0.22\% in an ideal
Gaussian distribution, leading to 19~MJ of stored energy for nominal
LHC parameters and 34~MJ in the case of
HL-LHC~\cite{helreview_valentino}. This leads to the conclusion that a
mechanism is needed to deplete the beam tails in a controlled
manner. Further information on the needs for halo control in LHC can
be found in Ref.~\cite{helreview}.

@@

\item \textbf{resonant excitation:} The electron beam is switched on
  only every \kth\ turn. The excitation can be represented
  by the following expression:
  \begin{eqnarray}
    \label{intro:eqn:1}
    f(t) & = & \sum_{n=-\infty}^{+\infty}\delta\left[t-n\cdot(kT)\right],
  \end{eqnarray}
  where $n$ is the turn number and $T$ is the revolution period. Its
  Fourier representation is
  \begin{equation}
    \label{intro:eqn:2}
    f(t) = \Sh_{kT}(t) = \frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt},
  \end{equation}
  with $f_n = n \cdot f_{\rm rev} / k$ and where $\Sh_{kT}$ is the
  Dirac comb. In general, \kthtp\ drives \kth-order
  resonances~\cite{md_sim_hel_res_ex_fitterer}. This type of pulsing
  pattern was used in the Tevatron during regular collider operations
  for abort-gap cleaning~\cite{hel_tevatron_abortgap_zhang}.
\end{itemize}

@@

\begin{figure*}
  \begin{tabular}{ccc}
    \multicolumn{3}{c}{\emph{Transverse damper off}} \\
    Random excitation, V (2017) & \seventhtp, V (2017) & \tenthtp, V
                                                         (2016) \\
    \includegraphics[width=\thirdwidth]{2017_emitv_avg_rel_vran_no_damper.png}&
    \includegraphics[width=\thirdwidth]{2017_emitv_avg_rel_v7th_no_damper_no_text.png} &
    \includegraphics[width=\thirdwidth]{2016_emitv_avg_rel_v10th_no_damper_no_text.png}\\
    \multicolumn{3}{c}{\emph{Transverse damper active}} \\
    Random excitation, V (2017) & \seventhtp, V (2017) & \tenthtp, V
                                                         (2016) \\
    \includegraphics[width=\thirdwidth]{2017_emitv_avg_rel_vran_with_damper.png} &
    \includegraphics[width=\thirdwidth]{2017_emitv_avg_rel_v7th_with_damper_no_text.png} &
    \includegraphics[width=\thirdwidth]{2016_emitv_avg_rel_v10th_with_damper_no_text.png} \\
    \end{tabular}
    \caption{Comparison of measured vertical emittances for bunches
      with transverse damper off (top row) and with transverse damper
      on (bottom row), during random V excitations (left), \seventhtp\
      in V (center) and \tenthtp\ in V (right). (Because of a change
      in experimental setup, for the bunches represented by the yellow
      line in the bottom right plot, the excitation was zero instead
      of the maximum value.)}
    \label{fig:damp}
\end{figure*}

@@

These studies were motivated in part by the need to assess the effects
of a hollow electron lens (HEL), which is being proposed for active
halo control in HL-LHC. Hollow electron beam collimation was
demonstrated in the Tevatron in continuous
mode~\cite{hel_tevatron_stancari}, i.e., with the same electron beam
current acting on a given bunch every turn. In this case, the unwanted
residual kicks, which arise from profile asymmetries, injection and
extraction bends, solenoid field misalignments, etc. have a negligible
effect on the circulating beam. An electron lens that is pulsed
resonantly can have much stronger effects, and this fact was used in
the Tevatron for abort-gap
cleaning~\cite{hel_tevatron_abortgap_zhang}. The resonant-pulsing
capability (having different currents every turn for a given set of
bunches) will be incorporated in the HL-LHC hollow electron lens to
enhance its range of achievable removal rates. However, due to time
constraints, only a very limited set of experimental tests with
resonant kicks could be done in the Tevatron. More studies are
currently being planned using the electron lenses in the Relativistic
Heavy Ion Collider (RHIC) at Brookhaven National
Laboratory~\cite{Fischer:PRL:2015}.
